Relationships in the extremes and their influence on competition and coexistence

### 2022-07-12 paper/Paper.Rnw
changing inva/resi

%Intro to storage effects and why they might be influenced by ATAs
In this paper, we show how ATAs can also be important for competition and for the coexistence of
competing species. In particular, we show how ATAs can strongly influence one specific coexistence mechanism:
storage effects.
%It is straightforward to see, intuitively, why ATAs may influence one mechanism of coexistence in particular: the storage effect. 
%We review, conceptually, what the storage effect is \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effect allows competing species to fluctuate %over time while escaping extinction because storage effects provide 
in abundance without any going extinct by providing
a relative benefit to whichever species is currently rare; this is the characteristic that makes the storage effect a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effect can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance), that is stronger when the species is common than when it is rare. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. 
Third, a buffering mechanism %must exist whereby species limit 
which limits the influence of competition %in times 
when the environment is poor. 
Dormant stages such as seed banks are an example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, the storage effect can be quantified by measuring to what extent EC covariance contributes to differences between 
rare and common species growth rates \citep{Chesson_1994,Ellner_2016}. %invader and resident 
Thus ATAs may, in principle, influence storage effect simply because storage effect hinges on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effect for species $i$ is 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
%Though some readers may find it slightly confusing to simultaneously use the variable $E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard, so it would be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like all operators, $\E(\cdot)$ is rendered upright.

We now specify how to decompose storage effect into contributions due to ATAs and 
due to $EC$ correlation \emph{per se}. 
Storage effect for species $i$ was computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which 
%the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
$E$ and $C$
were rendered independent of each other,
for both rare and common species [see (\ref{eq:storage_effects_gen})]. 
Because %these components 
$E$ and $C$ 
were rendered completely independent in the modified model, 
storage effect quantifies the contribution of the whole $EC$ relationship, in all its aspects, 
to the rare/common species growth rate difference. 
But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between %the components 
$E$ and $C$
altered relative to the original distributions.

>>>>

%Intro to storage effects and why they might be influenced by ATAs
In this paper, we show how ATAs can also be important for competition and for the coexistence of
competing species. In particular, we show how ATAs can strongly influence one specific coexistence mechanism:
storage effects.
%It is straightforward to see, intuitively, why ATAs may influence one mechanism of coexistence in particular: the storage effect. 
%We review, conceptually, what the storage effect is \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effect allows competing species to fluctuate %over time while escaping extinction because storage effects provide 
in abundance without any going extinct by providing
a relative benefit to whichever species is currently rare; this is the characteristic that makes the storage effect a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effect can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance), that is stronger when the species is at stable-state abundance than when it is rare. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. 
Third, a buffering mechanism %must exist whereby species limit 
which limits the influence of competition %in times 
when the environment is poor. 
Dormant stages such as seed banks are an example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Critically, the storage effect can be quantified in invasion analysis by measuring to what extent EC covariance contributes to differences between 
invader (rare) and resident (stable-state) species growth rates \citep{Chesson_1994,Ellner_2016}. %invader and resident 
Thus ATAs may, in principle, influence storage effect simply because storage effect hinges on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between invader and resident species mean growth rates, when $i$ is the invader.
Specifically, storage effect for species $i$ is 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
%Though some readers may find it slightly confusing to simultaneously use the variable $E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard, so it would be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like all operators, $\E(\cdot)$ is rendered upright.

We now specify how to decompose storage effect into contributions due to ATAs and 
due to $EC$ correlation \emph{per se}. 
Storage effect for species $i$ was computed by comparing
differences between invader and resident species mean growth rates for the original model and for a 
modified model for which 
%the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
$E$ and $C$
were rendered independent of each other,
for both invader and resideent species [see (\ref{eq:storage_effects_gen})]. 
Because %these components 
$E$ and $C$ 
were rendered completely independent in the modified model, 
storage effect quantifies the contribution of the whole $EC$ relationship, in all its aspects, 
to the invader/resident growth rate difference. 
But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between %the components 
$E$ and $C$
altered relative to the original distributions.
